Deterministic Bayesian inference methods for the Naomi model

### 2023-04-16 src/docs_18-04-23-lab-group/18-04-23-lab-group.Rmd
Presentation only lacking two plots from results notebook now #25

* Naomi is a complicated spatio-temporal evidence synthesis model
* Used by countries to produce HIV estimates in a yearly process supported by UNAIDS
* Fast inference is important to allow for interactive review and development of estimates
* Inference for Naomi is currently conducted using Template Model Builder (`TMB`) [@kristensen2015tmb]

@@

* Write and implement an algorithm for fast approximate Bayesian inference using the Laplace approximation and quadrature for the Naomi model
* Use `TMB` for writing the model in C++ and implementing the Laplace approximation via automatic differentitation
* Use adaptive Gauss-Hermite quadrature to integrate the hyperparameters

@@

```{r, echo=FALSE, fig.cap="One of the best engineering schools in Canada", out.width = '90%'}
knitr::include_graphics("the-plan.png")
```

@@

* For Malawi, Naomi has 24 hyperparameters: too many for a dense grid
* Proposed solution use principal components analysis and keep only first $s < 24$ dimensions
* We will use $k = 3$ points per dimension

@@

```{r, echo=FALSE, message=FALSE, fig.cap="Illustration of what this looks like for the 2D example, keeping only the first principal component, $s = 1$."}
gg3 <- gg
mvQuad::rescale(gg3, m = mu, C = cov, dec.type = 2)

@@

```{r, echo=FALSE, fig.cap="This Scree plot suggests 10 or so dimensions is enough. We use $s = 8$.", out.width = '80%'}
knitr::include_graphics("depends/tv-plot.png")
```

@@

* Use test data from Malawi
* Run HMC for as long as it takes to get a good answer
* Compare output of `TMB` with `aghq`: which is more similar to HMC?
* There are many ways you might do this: I'll show you Kolmogorov-Smirnov test results

@@

@@

* Working on a paper "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" based on this work
  * Joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
* Code and notebooks for this project are at `athowes.github.io/elgm-inf`

>>>>

* Naomi is a complicated spatio-temporal evidence synthesis model
* Used by countries to produce HIV estimates in a yearly process supported by UNAIDS
* Fast inference is important to allow for interactive review and development of estimates
* Inference for Naomi is currently conducted using Template Model Builder (`TMB`) [@kristensen2016tmb]

@@

* Write and implement an algorithm for fast approximate Bayesian inference using the Laplace approximation and quadrature for the Naomi model
* Use `TMB` for writing the model in C++, and implementing the Laplace approximation via automatic differentitation
* Use adaptive Gauss-Hermite quadrature to integrate the hyperparameters

@@

```{r, echo=FALSE, fig.cap="I attended one of the best engineering schools in Canada", out.width = '90%'}
knitr::include_graphics("the-plan.png")
```

@@

* For Malawi, Naomi has 24 hyperparameters: too many for a dense grid
* One proposed solution is to use principal components analysis (PCA) and keep only the first $s < 24$ dimensions
* We will use $k = 3$ points per dimension ($k = 2$ doesn't include the mode)

@@

```{r, echo=FALSE, message=FALSE, fig.cap="Illustration of what PCA-AGHQ looks like for the 2D example, keeping only the first principal component, $s = 1$."}
gg4 <- gg
mvQuad::rescale(gg4, m = mu, C = cov, dec.type = 1, levels = c(3, 1))

add_points(plot0, gg4) +
  labs(size = "Weight", caption = "")
```

@@

```{r, echo=FALSE, fig.cap="This Scree plot suggests 10 or so dimensions is enough. We use $s = 8$. Avoiding long computation times is still important.", out.width = '80%'}
knitr::include_graphics("depends/tv-plot.png")
```

@@

* We run HMC for as long as it takes to get a good answer (around 3 days)
* Compare output of `TMB` with `aghq`: which is more similar to HMC?
* There are many ways you might compare posterior distributions
* We'll show Kolmogorov-Smirnov test results here

@@

* Developed an approximate Bayesian inference method more accurate than `TMB` and faster than MCMC
* Scaled hyperparameter grid to >20 dimesions using PCA
* Relatively straightforward implementation using the `aghq` R package

# Future work

* More comprehensive inference comparison, particularly of model outputs
  * Maximum mean discrepancy
  * Pareto-smoothed importance sampling
* Extension to Laplace marginals via @wood2020simplified
  * May be more accurate, but have to think about computational costs

@@

* Working on a paper based on this work called "Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model" joint with Alex Stringer (Waterloo), Seth Flaxman (Oxford), Jeff Eaton (Imperial)
* Code and notebooks for this project are at `athowes.github.io/elgm-inf`
